---
title: "Check gene symbols against Entrez ID-s in **org.Hs.eg.db** database"
author: "Alexey Larionov"
date: "16 Feb 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
editor_options:
  chunk_output_type: console
---

# Summary

Explore gene names  

# Start section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
#options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors=F)

# Files and folders
base_folder="/Users/alexey/Documents/mg/s2021/zhang_tests"
working_folder=file.path(base_folder,"s04_explore_gene_list")
setwd(working_folder)

# This database package contains multiple gene-related ID-s including Entrez-ids and gene Symbols
#BiocManager::install("org.Hs.eg.db")
library(org.Hs.eg.db)

```

# Get all Entrez-ID-s & Symbols

```{r}

# Connect to the database with entrez ids and gene symbols
entrez_hs <- org.Hs.eg.db
entrez_hs
columns(entrez_hs)

# Select required data
#?select / help('select') / ?AnnotationDbi::select
entrez2symbols.df <- select(entrez_hs,
            keys=keys(entrez_hs,"ENTREZID"),
            columns=c("SYMBOL","ENTREZID"),
            keytype="ENTREZID")

head(entrez2symbols.df)

# Check missed data
sum(is.na(entrez2symbols.df))

# Clean-up
rm(entrez_hs)

```

# Read 166 genes

```{r}

data_file=file.path(base_folder,"d04_gene_lists","rms_genes_166.txt")
genes_166 <- scan(data_file,what="character")
genes_166

rm(data_file)

```

## Some missed genes

Some symbols from our list are missed in the SYMBOL field in **org.Hs.eg.db** database.  

This couls happen because  

- our list uses **Alias** instead of the current official Symbol  
- there is a **typo** in the symbol in our list  

```{r}

# Count genes from our list are present in the SYMBOL-s
sum(genes_166 %in% entrez2symbols.df$SYMBOL)

# Show genes from our list are absent in the SYMBOL-s
genes_166[! genes_166 %in% entrez2symbols.df$SYMBOL]

```

## No duplicated symbols

No symbols **from our list** are duplicated in the SYMBOL field in **org.Hs.eg.db** database  

```{r}

# A small number of Symbols is duplicated in **org.Hs.eg.db** database
sum(duplicated(entrez2symbols.df$SYMBOL))

# None of Symbols for our genes is duplicaed
x <- entrez2symbols.df$SYMBOL %in% genes_166
y <- entrez2symbols.df$SYMBOL[x]
sum(duplicated(y))

# Clean-up
rm(x,y)

```

# Save data

```{r}

save.image("s01_explore_gene_names.RData")

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
