---
title: "Test function to combine two GRanges"
author: "Alexey Larionov"
date: "25 Feb 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary
- Load functions  
    + *preprocess_sample.rms()*  
    + *combine_two_GRanges.rms()*  
- Toy data  
    + Read two Zheng samples to DataFrames  
    + Convert DataFrames to GRanges (using the same Seqinfo as in the 655 genes)  
- Preprocess  
    + Read 655 genes with coordinates (already in GRanges using Seqinfo(genome="hg38"))
    + Preprocess using *preprocess_sample.rms()*  
- Merge  
    + Use *combine_two_GRanges.rms()* function  

# Start section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors=F)

# Files and folders
base_folder <- "/Users/alexey/Documents/mg/s2021/zhang_tests"
scripts_fiolder <- file.path(base_folder,"scripts","s07_functions")
setwd(scripts_fiolder)

# Load functions
source(file.path(scripts_fiolder,"preprocess_sample.R"))
source(file.path(scripts_fiolder,"combine_two_GRanges.R"))

# Libraries: loaded when sourcing combine_two_GRanges.rms function
#library(GenomicRanges)
#library(SummarizedExperiment)

```

# Get toy data

## Sample 1 GRange

```{r}

# Sample data
sample1_name <- "RMSzhang_SJRHB020"
sample1_study <- "Zheng"

# Source file
sample1_file <- paste(sample1_name,"_hg38.bwa.QC.scores.filter.txt",sep="")
sample1_file_with_path <- file.path(base_folder,"data","d03_germline_annotated_files",sample1_file)

# Read to data frame
sample1.df <- read.table(sample1_file_with_path,
                         header=T, sep = "\t", quote = "", na.strings="-9")

# Coerce data frame to GRanges
sample1.gr <- GRanges(sample1.df, seqinfo=Seqinfo(genome="hg38"))

# Check result
summary(sample1.gr)
seqinfo(sample1.gr)
sort(sample1.gr)[,1:2]

# Clean-up (keep sample name)
rm(sample1_file_with_path,sample1.df)

```

## Sample 2 GRange

```{r}

# Sample data
sample2_name <- "RMSzhang_SJRHB022"
sample2_study <- "Zheng"

# Source file
sample2_file <- paste(sample2_name,"_hg38.bwa.QC.scores.filter.txt",sep="")
sample2_file_with_path <- file.path(base_folder,"data","d03_germline_annotated_files",sample2_file)

# Read to data frame
sample2.df <- read.table(sample2_file_with_path,
                         header=T, sep = "\t", quote = "", na.strings="-9")

# Coerse to GRanges
sample2.gr <- GRanges(sample2.df, seqinfo=Seqinfo(genome="hg38"))

# Check results
summary(sample2.gr)
seqinfo(sample2.gr)
sort(sample2.gr)[,1:2]

# Clean-up (keep sample name etc)
rm(sample2_file_with_path,sample2.df)

```

## Make samples dataframe

```{r}

# Make the data frame
samples_data.df <- data.frame(id=c(sample1_name,sample2_name),
                              file=c(sample1_file,sample2_file),
                              study=c(sample1_study,sample2_study))

rownames(samples_data.df) <- samples_data.df$id

samples_data.df

# Clean-up
rm(sample1_name,sample2_name,
   sample1_file,sample2_file,
   sample1_study,sample2_study)

```

## Read 655 (of 667) genes

```{r}

# Load genes coordinates
load(file.path(base_folder,"scripts","s04_gene_list_with_coordinates","s03_667_gene_coordinates.RData"))

# Remove unnecessary objects
rm(working_folder,absent_genes_9.df,genes_628.gr,genes_658.grl,genes_667.df,
   multi_genes_3.df, multi_genes_3.grl,multi_genes_30.df,multi_genes_30.grl)

# Update folders
base_folder <- "/Users/alexey/Documents/mg/s2021/zhang_tests"
scripts_fiolder <- file.path(base_folder,"scripts","s07_functions")

# Check the genes list
summary(genes_655.gr)
seqinfo(genes_655.gr)
genes_655.gr

```

# Preprocess samples

- keep variants on standard chromosomes  
- keep variants overlapping with the genes  
- exclude known common variants (known gnomAD MAF > 0.05)  

```{r}

# Preprocess
sample1.gr <- preprocess_sample.rms(sample1.gr, genes_655.gr)
sample2.gr <- preprocess_sample.rms(sample2.gr, genes_655.gr)

# Check result
summary(sample1.gr)
summary(sample2.gr)

# Clean-up
rm(genes_655.gr)

```

# Combine samples

Combines two GRanges to one SummarizedExperiment  
following the procedure agreed for RMS project  

```{r}

# Combine
merged.se <- combine_two_GRanges.rms(sample1.gr,sample2.gr,samples_data.df)

# Check result
merged.se

# Clean-up
rm(sample1.gr, sample2.gr, samples_data.df, combine_two_GRanges.rms)
# preprocess_sample.rms will be needed in the next test

```

# Save data

```{r}

save.image("s01_test_functions.RData")

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
