---
title: "Explore linking Annovar to VCF"
author: "Alexey Larionov"
date: "12 Feb 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
editor_options:
  chunk_output_type: console
---

# Summary

Explore linking Annovar-coded variants to mulrisample VCF  

# Start section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors=F)

# Folders
working_folder <- "/Users/alexey/Documents/mg/s2021/RMSzhang_SJRHB020"
setwd(working_folder)

# --- Library for reading VCF-s --- #

library(vcfR)

# --- Library and data-package(s) for ref genome --- #

#BiocManager::install("BSgenome")
#BiocManager::install("BSgenome.Hsapiens.NCBI.GRCh38")
# NCBI uses chromosome names w/o "chr" prefix
#BiocManager::install("BSgenome.Hsapiens.UCSC.hg38")
# UCSC uses chromosome names with "chr" prefix
#available.genomes()
#installed.genomes()

library(BSgenome)

# --- Other libraries --- #

#library(dplyr) # May interfere with many genomic packages !!!

```

# Read data

## Annotated text file

```{r}

SJRHB020.df <- read.table("RMSzhang_SJRHB020_hg38.bwa.QC.scores.filter.txt",
                         header=T, sep = "\t", quote = "", na.strings="-9")

dim(SJRHB020.df)
colnames(SJRHB020.df)

```

## Multisample VCF

```{r}

# Read vcf to vcfR object 
multisample.vcfr <- read.vcfR("zhang_hg38.bwa.QC.split_MA.vcf", verbose=F) # verbose=F

# Check result
multisample.vcfr
#head(multisample.vcfr)

# Get list with data from VCF header and fixed/standard VCF columns
meta_fix <- vcfR2tidy(multisample.vcfr, info_only=T)

# Get data frame with variants information
variants.df <- as.data.frame(meta_fix$fix)
dim(variants.df)
colnames(variants.df) 

# Make unique key for each variant
key <- paste(variants.df$CHROM, variants.df$POS, variants.df$REF, variants.df$ALT, sep="_")
key[1:5]

# Add key to variants data frame
variants.df <- data.frame(key,variants.df)
key -> rownames(variants.df)
variants.df[1:5,1:6]

# Clean-up
rm(meta_fix, multisample.vcfr, key)

```

## Reference genome

Using a BSgenome data-package  

Alternatively the genome could be red directly from a FASTA file:  
https://www.biostars.org/p/171577/  

```{r}

# Check available and installed BSgenome data packages
# available.genomes()
#installed.genomes()
# NCBI uses chromosome names w/o "chr" prefix
# UCSC uses chromosome names with "chr" prefix

# Load required genome: b38 with "chr" prefixes
#b38 <- getBSgenome("BSgenome.Hsapiens.NCBI.GRCh38")
b38 <- getBSgenome("BSgenome.Hsapiens.UCSC.hg38")

# Sequence / chromosome names
b38@seqinfo

# Example of bases retrieval by coordinates
getSeq(b38,"chr1",start=257894,end=257897)

```

# Linking variants  

## SNP

- VCF.Chr = Annovar.Chr  
- VCF.Pos = Annovar.Start  
- VCF.Ref = Annovar.Ref  
- VCF.Alt = Annovar.Alt  

```{r}

# Example of SNP in Annotated text file
annovar.snp <- SJRHB020.df[32,c("Chr","Start","End","Ref","Alt")]
annovar.snp

# Checking SNP position in b38 ref genome
#getSeq(b38,"chr2",start=197421031,end=197421031)

# Translating SNP coordinates from annovar to vcf
vcf.snp.chr <- annovar.snp$Chr
vcf.snp.pos <- annovar.snp$Start
vcf.snp.ref <- annovar.snp$Ref
vcf.snp.alt <- annovar.snp$Alt
vcf.snp <- c(vcf.snp.chr, 
             as.character(vcf.snp.pos), 
             vcf.snp.ref, 
             vcf.snp.alt)

# Make the key to search multisample VCF
vcf_key <- paste(vcf.snp, collapse="_")
vcf_key

# Check this SNP in Multisample VCF
if(vcf_key %in% variants.df$key){
  variants.df[vcf_key,c("CHROM","POS","REF","ALT")]
} else {
  cat("Can't find variant in multisample VCF\n")
}

# Clean-up
rm(annovar.snp,vcf.snp,vcf.snp.chr,vcf.snp.pos,vcf.snp.ref,vcf.snp.alt,vcf_key)

```

## Deletion

- VCF.Chr = Annovar.Chr  
- VCF.Pos = Annovar.Start-1 (because VCF adds one ancoring nucleotide)  
- VCF.Ref = Ref.genome[VCF.Pos, Annovar.End] (sequence from ref genome)  
- VC.Alt = Ref.genome[VCF.Pos] (ancoring nucleotide from ref genome)  

```{r}

# Example of DEL in Annotated text file
annovar.del <- SJRHB020.df[1,c("Chr","Start","End","Ref","Alt")]
annovar.del

# Checking DEL position in b38 ref genome
#getSeq(b38,"chr19",start=52300417,end=52300418)

# Translating DEL coordinates from annovar to vcf
vcf.del.chr <- annovar.del$Chr
vcf.del.pos <- annovar.del$Start - 1

vcf.del.ref <- getSeq(b38,vcf.del.chr,
                      start=vcf.del.pos,
                      end=annovar.del$End)

vcf.del.alt <- getSeq(b38,vcf.del.chr,
                      start=vcf.del.pos,
                      end=vcf.del.pos)

vcf.del <- c(vcf.del.chr, 
             as.character(vcf.del.pos), 
             as.character(vcf.del.ref), 
             as.character(vcf.del.alt))
vcf.del

# Make the key to search multisample VCF
vcf_key <- paste(vcf.del, collapse="_")
vcf_key

# Check this DEL in Multisample VCF
if(vcf_key %in% variants.df$key){
  variants.df[vcf_key,c("CHROM","POS","REF","ALT")]
} else {
  cat("Can't find variant in multisample VCF\n")
}

# Clean-up
rm(annovar.del,vcf.del,vcf.del.chr,vcf.del.pos,vcf.del.ref,vcf.del.alt,vcf_key)

```

## Insertion

- VCF.Chr = Annovar.Chr  
- VCF.Pos = Annovar.Start
- VCF.Ref = Ref.genome[Annovar.Start]  
- VC.Alt = VCF.Ref + Annovar.Alt

```{r}

# Example of INS in Annotated text file
annovar.ins <- SJRHB020.df[6,c("Chr","Start","End","Ref","Alt")]
annovar.ins

# Checking INS position in b38 ref genome
#getSeq(b38,"chr12",start=51346631,end=51346631)

# Translating INS coordinates from annovar to vcf
vcf.ins.chr <- annovar.ins$Chr
vcf.ins.pos <- annovar.ins$Start

vcf.ins.ref <- getSeq(b38,vcf.ins.chr,
                      start=vcf.ins.pos,
                      end=vcf.ins.pos)

vcf.ins.alt <- paste0(vcf.ins.ref, annovar.ins$Alt, collapse="")

vcf.ins <- c(vcf.ins.chr, 
             as.character(vcf.ins.pos), 
             as.character(vcf.ins.ref), 
             as.character(vcf.ins.alt))
vcf.ins

# Make the key to search multisample VCF
vcf_key <- paste(vcf.ins, collapse="_")
vcf_key

# Check this INS in Multisample VCF
if(vcf_key %in% variants.df$key){
  variants.df[vcf_key,c("CHROM","POS","REF","ALT")]
} else {
  cat("Can't find variant in multisample VCF\n")
}

# Clean-up
rm(annovar.ins,vcf.ins,vcf.ins.chr,vcf.ins.pos,vcf.ins.ref,vcf.ins.alt,vcf_key)

```

# Count types of variants

Count SNP, DEL and INS variants. Check if any variants fall out of these categories: so they should be linked manually.  

```{r}

# Count SNP/SNV-s
SNP <- SJRHB020.df$Ref!="-" & SJRHB020.df$Alt!="-" & 
  nchar(SJRHB020.df$Ref)==1 & nchar(SJRHB020.df$Alt)==1
cat("There are",sum(SNP),"SNP-s\n")

# Count Insertions
INS <- SJRHB020.df$Ref=="-" & SJRHB020.df$Alt!="-"
cat("There are",sum(INS),"insertions\n")

# Count Deletions
DEL <- SJRHB020.df$Ref!="-" & SJRHB020.df$Alt=="-"
cat("There are",sum(DEL),"deletions\n")

# Count remained non-translatable variants
# (such as multi-nucleotide variants, e.g. AT -> CGG)
MNV <- !SNP & !INS & !DEL

if(sum(MNV)>0){
  cat(sum(MNV),"variants renain to be translated manually\n")
} else {
  cat("There is no remaining non-translated variants\n")
}

# Check for absurd -/- variants
ABSV <- sum(SJRHB020.df$Ref=="-" & SJRHB020.df$Alt=="-")

if(sum(ABSV)>0){
  cat("There are",ABSV,"absurd -/- variants\n")
} else {
  cat("There is no absurd -/- variants\n")
}

# Cleran-up
rm(ABSV,SNP,INS,DEL,MNV)

```

# Save data

```{r}

save.image("s04_linking_Annovar_to_VCF.RData")

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
