---
title: "Get coordinates of 667 genes"
author: "Alexey Larionov"
date: "20 Feb 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
editor_options:
  chunk_output_type: console
---

# Summary

Script selects coordinates for 655 out of 667 genes: **655 = 667 - 9 - 3**  

9 Genes are not present in the transcripts database because they are not transcribed or missed Entrez ID.  

30 genes have multiple locations in Ref genome.  

27 of these genes have the "main" location in standard chromosomes and "additional" location(s) in Alt loci.  For such genes only coordinaltes on standard chromosomes are taken.  

In 3 genes the multiple locations were reported within standard chromosomes:  

- P2RY8 (286530) X and Y  
- CRLF2 (64109) X and Y  
- SSX2 (6757) X+ and X-  

Coordinates for such genes could be dealt with on an ad-hoc basis.  In this script these 3 genes are just excluded from the final list.  

# Start section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors=F)

# Files and folders
base_folder="/Users/alexey/Documents/mg/s2021/zhang_tests"
working_folder=file.path(base_folder,"s04_get_gene_coordinates")
setwd(working_folder)

# Libraries
library(dplyr)

```

# Transcripts database

This database package is used to retrieve coordinates of human transcripts; also it provides exones and CDS; the genes are given by entrez-ids  

```{r}

#BiocManager::install("TxDb.Hsapiens.UCSC.hg38.knownGene")
library(TxDb.Hsapiens.UCSC.hg38.knownGene)

# Connect to the database
b38.tx <- TxDb.Hsapiens.UCSC.hg38.knownGene

# Check connection
b38.tx
length(seqlevels(b38.tx))
seqlevels(b38.tx)[1:30]
columns(b38.tx)

# This command can be used to keep only standard chromosomes
# (i.e. to remove Alt loci)
#keepStandardChromosomes(b38.tx)

```

# Read 667 genes

```{r}

# Load data
load(file.path(working_folder,"s02_667_gene_ids.RData"))

# Update folders (just in case)
base_folder="/Users/alexey/Documents/mg/s2021/zhang_tests"
working_folder=file.path(base_folder,"s04_get_gene_coordinates")

```

# Explore transcripts database

## 628 genes with a single location in Ref genome

Function *genes()* retrieves gene locations (as GRanges) from the Transcript Database: see *?genes*.

The default *genes()* setting **single.strand.genes.only=T** does not include genes with multiple locations in **reference** genome, such as  

- located in canonical and in Alt loci  
- duplicated in X and Y chromosomes (e.g. in pseudoautosomal regions)  
- some other cases, as will be seen below  

Finaly, some "genes" that have Entrez ID-s may have no transcripts, such as  

- pseudogenes  
- loci that have combinatorial transcription, e.g. receptors and molecules related to immune responce  

Thus, some genes included in 667 list are not retrieved by the default *genes()* call.   

```{r}

genes_628.gr <- genes(b38.tx, 
                      filter=list(gene_id = genes_667.df$ENTREZID))

length(genes_628.gr)

genes_628.gr

```

## 30 genes with multiple locations in Ref genome

To show all genes present in the database, the dafult *genes()* settings should be overwritten by **single.strand.genes.only=F**.  

Importantly, the output will change from *genomic ranges* to *genomic ranges* **list**: one *geniomic ranges* obgect per gene, where some genes are presented by several ranges.  

```{r}

# Get genes from our list
genes_658.grl <- genes(b38.tx, 
                       filter=list(gene_id = genes_667.df$ENTREZID), 
                       single.strand.genes.only=F)

# Check the number of genes
length(genes_658.grl)

# Count genes by the locations number in Ref genome
table(elementNROWS(genes_658.grl))

# Note the convinience function elementNROWS() for GRL
# It is equivalent to unlist(lapply(grl,length))
# Its contrintuitive that length(grl) somehow outputs list instead of vector
#table(unlist(lapply(genes_658.grl,length))) 

# Genes with multiple locations in Ref genome
multi_genes_30.grl <- genes_658.grl[elementNROWS(genes_658.grl)>1]
length(multi_genes_30.grl)
multi_genes_30.grl

# Make the dataframe to add Gene names
multi_genes_30.df <- data.frame(ENTREZID=names(multi_genes_30.grl),
                                n_loci=elementNROWS(multi_genes_30.grl))

multi_genes_30.df <- left_join(multi_genes_30.df,genes_667.df)
multi_genes_30.df %>% arrange(desc(n_loci))

```

## 3 Genes with multiple locations on standard chromosomes

P2RY8 (286530) X and Y
CRLF2 (64109) X and Y
SSX2 (6757) X+ and X- 

```{r}

# Remove Alt loci from transcript database: keep only standard chromosomes
# Note using keepStandardChromosomes() function that is available for 
# Annortation-databases instead of my previous manual handling of this step
keepStandardChromosomes(b38.tx)
seqlevels(b38.tx)

# Get genes from 667 list with transcripts on standard chromosomes
genes_658.grl <- genes(b38.tx, 
                       filter=list(gene_id = genes_667.df$ENTREZID), 
                       single.strand.genes.only=F)
length(genes_658.grl)

# Genes with multiple locations on standard chromosomes
multi_genes_3.grl <- genes_658.grl[elementNROWS(genes_658.grl)>1]
length(multi_genes_3.grl)
multi_genes_3.grl

# Make data frame with the gene names
# note function stak(), which is specific for GRL: it converts GRL to GR
multi_genes_3.df <- left_join(data.frame(stack(multi_genes_3.grl)),genes_667.df,
                              by = c("gene_id"="ENTREZID"))
multi_genes_3.df

```

## 9 genes absent in the transcripts database 

The first gene may be recorded as FGFR1OP
The second is a pseudogene?
The remaining "genes" are loci with combinatorial transcription, which might be absent in a standard transcript database, e.g. :

- https://www.ncbi.nlm.nih.gov/gene/3492  
- https://www.ncbi.nlm.nih.gov/gene/6955  

```{r}

absent_genes_9.df <- genes_667.df[!genes_667.df$ENTREZID %in% names(genes_658.grl),]

dim(absent_genes_9.df)

absent_genes_9.df

```

# 655 genes list  

Includes  

- 628 genes with unique locations on standard chromosomes  
- 27 genes with multiple locations, where only one of them is on standard chromosomes  

```{r}

# Alt loci have already been removed from transcript database 
#keepStandardChromosomes(b38.tx)
seqlevels(b38.tx)

# Use default genes() setting to get rid of genes with multiple locations in genome
genes_655.gr <- genes(b38.tx, 
                      filter=list(gene_id = genes_667.df$ENTREZID))
length(genes_655.gr)

# Sort by position
genes_655.gr <- sort(genes_655.gr)

# Check result
genes_655.gr

```

# Add symbols

```{r}

# Prepare data frame with gene names for 655 GR: 
# select by Entrez ID and sync the order with GR
genes_655.df <- genes_667.df[genes_667.df$ENTREZID %in% genes_655.gr$gene_id,]
rownames(genes_655.df) <- genes_655.df$ENTREZID
genes_655.df <- genes_655.df[genes_655.gr$gene_id,]

# Check result
head(genes_655.gr)
head(genes_655.df)

# Add Symbols to GR object
genes_655.gr$symbol <- genes_655.df$SYMBOL

# Check result
genes_655.gr

# Clean-up
rm(genes_655.df)

```

# Save data

```{r}

save.image("s03_667_gene_coordinates.RData")

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
